Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene

### 2023-02-05 manuscript/manuscript.Rmd
Update manuscript.Rmd

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. nearest living relative...

We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.

>>>>

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. nearest living relative...

We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012]. The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modellers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. More recent climate models have been able to align their simulations with the proxy record [@Huber2013;...]. A comparison of multiple proxy-based early Eocene reconstructions of global average temperatures @Inglis2020

Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
